Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Remove comments from Methods section, rewrite first part of Results section to correspond to the newly formulated simulation objectives.

<!-- Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

@@

This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2]. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred. Yet, monitoring convergence of iterative imputation algorithms diagnostically is challenging. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Therefore, the aim of convergence diagnostics for MCMC methods is not to establish the point at which convergence is reached, but to monitor signs of non-convergence [@hoff09]. Several of such diagnostics exist for MCMC methods, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. The convergence properties of the MICE algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.

**Include sub-questions!** How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?** *(Maybe add these too? What are the effects of continued iterations on estimates, predictions and inferences? Do the answers differ with varying missingness proportions? That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)*

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. **The number of units $i$ with at least one missing data value can be divided by the total number of units $n$ to obtain the 'missingness proportion'  $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain). **

@@

The goal of imputation through e.g. FCS/MICE is to pbtain a multivariate distribution by iterating over the sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the Rhat and autocorrelation. The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of  evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it....

@@

**In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again, now together---as are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up **in the typical convergence situation**. The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$. Panel C displays the $AC$ as computed with the R function `stats::acf()`. When we look at this panel, we conclude something weird. The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on stats::acf() [@R], we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to @box15, the function works when the number of iterations > 50. The default number of iterations in iterative imputation, however, is often much lower. Therefore, we compute $AC$ manually, see panel D. The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$. Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence. And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice. That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.

@@

@@

```
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes across simulation runs

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14]. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem **(from an unspecified social scientific field of study)**. Let the predictor space be defined as 
$$
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
$$
A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome $Y$. For each unit $i = 1, 2,..., N$, let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of each quantity of scientific interest $Q$.

@@

We consider four types of $Q$s that are often of interest in empirical research. Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j = Y, X_1, X_2, X_3$. We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
where $Y'$ is the expected value of the outcome. And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**. The true values of these scientific estimands are calculated with the R functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R]. Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).

@@

The complete dataset is amputed with the `mice` package [function `mice::ampute()`; @mice]. All missingness is multivariate, and conform a 'missing completely at random' missingness mechanism [@rubin87]. I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion ($p_{mis} =.05,.25,.5,.75,.95$). 
<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87]. The probability to be missing is the same for all $n \times k$ cells in $y$ [a 'missing completely at random' missingness mechanism; @rubin87]. -->

Missing datapoints in $y$ are imputed with the `mice()` function [@mice]. All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($M=5$). The number of iterations varies between simulation conditions ($T=1,2,...,100$). 
<!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$). From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).  -->

After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}, where $m = 1, 2, ..., 5$.

- on the 5 imputations $y_{imp,m}$: get mean and var within imp chains as theta to compute conv diag --> then complete with `mice::complete()`

- on the 5 completed datasets {$y_{obs}, y_{imp, m}$}: 1) estimate Qhats for mu and sigma --> pool as mean() --> get Qbars of mu and sigma; 2) compute lambda as theta for conv diag; 3) perform lm to get Qhats for betas and r^2 --> use Qhats of beta as theta to compute conv diag --> pool Qhats of beta and r^2 to get Qbars

The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}$s across imputations. To calculate the $\bar{Q}$s, we combine the observed data $y_{obs}$ and the imputed data for each imputation $y_{imp,m}$ with the function `mice::complete()`. Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$). Estimated regression coefficients are obtained with the function `stats::lm()` for each imputation, and then pooled conform @vink14. The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.

For each run of the imputation algorithm, we compute non-convergence diagnostics $\widehat{R}$ and $AC$. We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration. The two methods are applied to four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), a statistic of scientific interest $\hat{Q}$ in each {$y_{obs}, y_{imp, m}$}, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.

The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates. Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$. Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**). **remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**

@@

The impact of inducing non-convergence in the iterative imputation algorithm is assessed by comparing $\bar{Q}$s with $Q$s. For each $Q$, we compute bias as $\bar{Q} - Q$. For the regression coefficients $Q=\beta_{1,2,3}$, we also compute the empirical coverage rate (CR). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Let
$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. Since under-estimating the variance of $\bar{Q}$ may yield spurious inferences, we also inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.

The performance of two methods is evaluated against the bias and confidence validity of the estimated $Q$s.

@@

We want to know what the effect is of inducing non-conv of the bias in all Qs and the confidence validity of the betas. So we plot the bias for each Q against the number of iterations and the missingness proportion in the condition. We look at averages across simulation repetitions. Is there more bias in conditions where non-conv was induced? (NOPE for univariates, YES for multivar.).

Then, we want to know if conditions in which the Qs are affected by non-convergence (on average) are also identified by the conv diag. We do this by looking at the bias in Q against the conv diag applied to related thetas. Is the point at which there is no more bias the same point at which the conv diag indicate minimal signs of non-conv? (NOPE, they are too conservative).

Is one of the thetas better? (NAH, doesn't really matter).

What should the thresholds be? (LOWER than recommended) And the default nr of iterarions? (MAYBE increase to 10?)

**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset. 3) Add more info about figure legends and axes.)**

For reasons of brevity, we only discuss the convergence diagnostics for the $Q$ with the worst performance in terms of bias. For the descriptve statistics, the magnitude of the bias was the largest in $Y$ (i.e, $j = Y$ in $\mu_j$ and $\sigma_j$). For the $Q=\beta_{1,2,3}$, $\beta_1$ showed the worst perfomance (i.e, the effect of $X_1$ on $Y$. Moreover, we only show conditions where $T\leq50$ in the figures. Full results are available from **[link github repo]**.

@@

As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s. Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases. The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ. This means that the algorithm has converged sufficiently (**under the current specifications**).

As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures. $\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data. Autocorrelation values were indeed decreasing with a higher number of iterations.

@@

**Message: after one iteration (T=2), the algorithm is as well converged as the missingness proportion allows. Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** Number of it has little influence on bias in hat{Q}, only in first iteration some bias in missingness conditions of 25 and up. There is a clear different between the missingness conditions. The higher the missingness prop, the larger the magnitude of the bias. For this Q, t=2 would be sufficient. And only for p_mis = 5 the bar{Q}s are unbiased. All other p_mis conditions have bias irrespective of the number of iterations.

>>>>

<!-- $Q$ is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

@@

This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2]. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---yet not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics we may only identify signs of non-convergence [@hoff09]. Several of such diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. The convergence properties of the MICE algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. **add societal relevance: what is the consequence of non-convergence?**

**Include sub-questions and add structure of the paper!** How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?** *(Maybe add these too? What are the effects of continued iterations on estimates, predictions and inferences? Do the answers differ with varying missingness proportions? That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)*

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. **The number of units $i$ in dataset $y$ with at least one missing data value, divided by the total number of units $n$, is called the missingness proportion in $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain). **

@@

**The goal of imputation through e.g. MICE is to obtain a multivariate distribution by iterating over the sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the $\widehat{R}$ and autocorrelation. The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it.... **

@@

**In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again, now together---as are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up **in the typical convergence situation**. The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$. Panel C displays the $AC$ as computed with the R function `stats::acf()`. When we look at this panel, we conclude something weird. The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on stats::acf() [@R], we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to @box15, the function works when the number of iterations > 50. The default number of iterations in iterative imputation, however, is often much lower. Therefore, we compute $AC$ manually, see panel D. The $AC$ values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$. Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence (**add thetas**). And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice. That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$. (**add to formulate advice on thetas?**)

@@

5. **Something about thetas**

@@

```
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes of all simulation runs

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14]. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem **(from an unspecified social scientific field of study)**. Let the predictor space be defined as 
$$
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
$$
A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome $Y$. For each unit $i = 1, 2,..., N$, let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s. **MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?** $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ $\hat{Q}$s to obtain a $\bar{Q}$ for each $Q$. **Add that we try to predict the effect of non-converegence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by appliying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**

@@

We consider four types of $Q$s that are often of interest in empirical research. Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j = Y, X_1, X_2, X_3$. We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
where $Y'$ is the expected value of the outcome. And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**. The true values of these scientific estimands are calculated on the complete data with the R functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R]. 
<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).  -->

@@

In each simulation repetition, we ampute the complete dataset five times to obtain five different incomplete datasets, $y$s. The ratio between $y_{obs}$ and $y_{mis}$ in each $y$ depends on the missingness proportion: $p_{mis} =.05,.25,.5,.75,.95$. We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice]. All missingness is multivariate, and conform a 'missing completely at random' missingness mechanism [@rubin87]. I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{mis}$. 
<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87]. The probability to be missing is the same for all $n \times k$ cells in $y$ [a 'missing completely at random' missingness mechanism; @rubin87]. -->

Missing datapoints in the incomplete datasets, $y$s, are imputed with the `mice()` function [@mice]. All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($M=5$). Because $M=5$, each run of the `mice()` function results in five sets of imputations, $y_{imp,m}$s, where $m = 1, 2, ..., 5$. The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$). 
<!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$). From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).  -->

From the 5 imputations $y_{imp,m}$: get chain means and chain variances from imputation object to use as scalar summary $\theta$ in convergence diagnostics. Then use `mice::complete()` to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.

On the 5 completed datasets {$y_{obs}, y_{imp, m}$}, we:

- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. Then pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.

- Compute $\lambda_1$, to use as $\theta$ in convergence diagnostics. By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_1$, is equal to the variance of the principal component solution of each {$y_{obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].

- Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}$s for $Q=\beta$ and $Q=r^2$. Pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ (**check  pooling of  $r^2$, instead of `mice::pool.r.squared()`**). Additionally, use the $\hat{Q}$s of $Q=\beta$ as $\theta$ to compute convergence diagnostics.

<!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.  -->

<!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}$s across imputations. To calculate the $\bar{Q}$s, we combine the observed data $y_{obs}$ and the imputed data for each imputation $y_{imp,m}$ with the function `mice::complete()`. Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$). Estimated regression coefficients are obtained with the function `stats::lm()` for each imputation, and then pooled conform @vink14. The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.  -->

We apply non-convergence diagnostics $\widehat{R}$ and $AC$ to each scalar summary $\theta$. We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.

@@

We assess the impact of inducing non-convergence in the iterative imputation algorithm by comparing $\bar{Q}$s with $Q$s. For each $Q$, we calculate bias as $\bar{Q} - Q$. For the regression coefficients $Q=\beta_{1,2,3}$, we also compute the empirical coverage rate (CR). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Let
$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are conficence valid estimates of the $Q$s. Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$. CIW is of interest because too short CIs indicate under-estimation of the variance in $\bar{Q}$, which may yield spurious inferences.

With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the $\bar{Q}$s are *not* unbiased, confidence valid estimates of the $Q$s. **Rephrase and add corresponding $Q$s:** The two methods are applied to four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), the estimated regression coefficients $\beta_{1,2,3}$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$), and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.

@@

**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset?? 3) Add more info about figure legends and axes.)**

Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s. Specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$. Therefore, we plot these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$). We look at averages across simulation repetitions.

Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergance diagnostics $\widehat{R}$ and $AC$. We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-the convergence in the $\theta$s.

- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).

- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics? (NOPE, they are too conservative).**

- Is one of the thetas better? (NAH, doesn't really matter).

- What should the thresholds be? (LOWER than recommended).

- And the default nr of iterarions? (MAYBE increase to 10?).

For reasons of brevity, we only discuss the convergence diagnostics for the $Q$s with the worst performance in terms of bias. Bias in the  descriptve statistics ($Q=\mu$ and $Q=\sigma$) was most pronounced in the outcome variable $Y$. Of the regression coefficients, $Q=\beta$, the largest bias was observed in the effect of $X_1$ on $Y$, $\beta_1$. Results for $Q=r^2$ are reported as-is. In the figures, however, we only present simulation conditions where $T\leq50$. Full results are available from **[link github repo]**.

@@

As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence. I.e., on average these conditions portet more extreme bias in the estimated $Q$s, and non-nominal coverages. Roughly speaking this means that MICE is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$. The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ. This means that the algorithm has converged sufficiently (**under the current specifications**).

As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures. **Add split between convergence diagnostics versus performance measures, and convergence diagnostics versus simulation conditions.** $\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data. Autocorrelation values were indeed decreasing with a higher number of iterations, and lower in conditions with less missingness.

@@

**Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows. Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** Number of it has little influence on bias in $\hat{Q}$, only in first iteration some bias in missingness conditions of 25 and up. There is a clear different between the missingness conditions. The higher the missingness prop, the larger the magnitude of the bias. For this $Q$, t=2 would be sufficient. And only for $p_{mis}$ = 5 the $\bar{Q}$s are unbiased. All other $p_{mis}$ conditions have bias irrespective of the number of iterations.

$\widehat{R}$ differs a lot across iteration conditions. But not structurally for missingness proportions. Highest $\widehat{R}$s for the condition $T$=2, while the most bias is for ($T$=1), but there is no $\widehat{R}$ for that by definition. But for missingness, 25% seems the most signs of non-conv, while it has one of the least bias. Threshold of 1.2 reached in condition it = 3, $p_{mis}$ = 5 and 75, and then for it>6, $p_{mis}$ =  5 and 50, and $T$>10 for all $p_{mis}$. Threshold 1.1 reached in in $T$=12, $p_{mis}$ = all but 75, 1.01 only reached in $T$ = 93, $p_{mis}$ = 75. While for this $Q$, $T$=2 would be sufficient.
